---
title: "Bayesian Detection of Differential Item Functioning Under the Rasch Model"
author: "Monika Bagociute"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    code_folding: hide
    toc: true
    toc_depth: 3
---

## Research overview

This project investigates the detection of Differential Item Functioning (DIF)
under the Rasch model using Bayesian estimation.

The analysis has two components. First, a simulation study evaluates whether a
custom Metropolis–Hastings sampler can recover known DIF parameters and compares
its estimates with those obtained from JAGS. Second, the Bayesian approach is
applied to an empirical assessment dataset and compared with the classical
Mantel–Haenszel procedure.

### Research question

Can Bayesian estimation reliably detect DIF under the Rasch model, and how do
the results compare with a classical DIF detection procedure?

### Analytical workflow

The analysis demonstrates:

- Simulation-based validation of a statistical estimation procedure
- Bayesian parameter estimation using Metropolis–Hastings MCMC
- Validation against JAGS
- Posterior means and credible intervals
- MCMC diagnostics, including trace and autocorrelation plots
- Empirical DIF analysis
- Comparison of Bayesian and Mantel–Haenszel results
- Bayesian hypothesis testing

### Tools

- R
- R Markdown
- JAGS
- Metropolis–Hastings MCMC
- Bayesian estimation
- Rasch modelling
- DIF analysis



#### Simulation 

Data were simulated under a Rasch model. The dataset consisted of 500 participants responding to 10 dichotomous items. Participants were randomly assigned to one of two groups, which may be interpreted, for example, as boys and girls. Differential Item Functioning (DIF) was introduced for Items 3 and 8 by allowing group membership to affect the probability of a correct response, whereas the remaining items were assumed to function equivalently across groups.

```{r data}



library(rjags)
library(coda)
#library(difR)
library(bain)
library(restriktor)
library(knitr)
library(xtable)

set.seed(123456)

# persons
N <- 500 
# items
J <- 10       

# assign them randomly to a group
group <- rbinom(n = N, size = 1, prob = .50)
table(group)

# draw the ability estimates from the normal distrib,
theta <- rnorm(n = N, mean = 0, sd = 1)
#theta
# draw the difficulty parameters from the normal distrib.
b <- rnorm(n = J, mean = 0, sd = 1)
#b

# Introducing DIF
delta <- rep(0, J)

# item 3 harder
delta[3] <- 1

# Item 8 easier 
delta[8] <- -1

#delta

```
We continue generating the responses.

```{r}

# response matrix (empti)
Y <- matrix(0, nrow = N, ncol = J) 
#Y

# generating responses
for(i in 1:N){
for(j in 1:J){
eta <-theta[i] -b[j] - delta[j] * group[i] 
p <- plogis(eta) # rasch function 
Y[i,j] <- rbinom(n = 1, size = 1, prob = p)
  }
}

# check up 
dim(Y) # correct dim

round(colMeans(Y), 3) # we see the % of responding correctly


rbind(
  Reference = colMeans(Y[group == 0, ]),
  Focal = colMeans(Y[group == 1, ])
) # for item 3 and item 8 we see a clear difference



```
The simulated data seems correct. 

```{r}

# storing
simdata <- list(Y = Y, theta_true = theta, b_true = b, delta_true = delta,
  group = group
)

saveRDS(
  simdata,
  "../Data/processed/simulated_rasch_dif.rds"
)
```


Theoretical curves using the generating mechanism. I am using this information for my final report only. 
```{r}

pdf("../Figures/theoretical_item_response_curves.pdf",
    width = 12,
    height = 7)

par(mfrow = c(2, 5))

theta_grid <- seq(-3, 3, length.out = 200)

for (j in 1:J) {

  p_ref <- plogis(theta_grid - b[j])

  p_focal <- plogis(
    theta_grid - b[j] - delta[j]
  )

  plot(
    theta_grid,
    p_ref,
    type = "l",
    ylim = c(0, 1),
    main = paste("Item", j),
    xlab = "Theta",
    ylab = "P"
  )

  lines(
    theta_grid,
    p_focal,
    lty = 2
  )
}

dev.off()
```


#### JAGS

Since the true values of the item and DIF parameters are known in the simulated data, agreement between the estimates obtained from the JAGS and Metropolis-Hastings sampler provides evidence that the implemented algorithm correctly approximates the posterior distributions.


```{r}


model_string <- "
model{

for(i in 1:N){

for(j in 1:J){

Y[i,j] ~ dbern(p[i,j])

logit(p[i,j]) <- theta[i] - b[j] - delta[j]*group[i]} }

  # Priors

for(j in 1:J){

b[j] ~ dnorm(0,1)

delta[j] ~ dnorm(0,1)}}
"

```


Data for JAGS

```{r}
library(rjags)

jags_data <- list(
Y = Y,
N = N,
J = J,
group = group, 
theta = theta

)


```

Initial values

```{r}
inits <- function(){
list(b = rnorm(J,0,1), delta = rep(0,J))
}

```

Fitting the model 

```{r}

mod <- jags.model(
  textConnection(model_string),
  data = jags_data,
  inits = inits,
  n.chains = 3,
  n.adapt = 2000
)

```

```{r}
# burn-in
update(mod, 5000) # takes some time

# draw posterior samples

library(coda)

samples <- coda.samples(model = mod, variable.names = c("b","delta"), n.iter = 10000)

```
Extract JAGS estimates
```{r}
summary(samples)

stats <- summary(samples)$statistics
delta_rows <- grep("^delta", rownames(stats))
delta_est <- stats[delta_rows,"Mean"]
delta_est


```




####  MH sampler 


```{r}

set.seed(123)


n_iter <- 10000
burnin <- 2000


prop_b <- 0.3
prop_delta <- 0.3


# Initial values
b_mh <- rep(0, J)
delta_mh <- rep(0, J)


# Storage

b_store <- matrix(NA, n_iter, J)
delta_store <- matrix(NA, n_iter, J)

# vectors count how many proposals are accepted for each item
acc_b <- rep(0, J)
acc_delta <- rep(0, J)
```

Log likelihood for a single item

```{r}

loglik_item <- function(b, delta, y, theta, group){
eta <- theta - b - delta*group
p <- plogis(eta)
sum(y*log(p) + (1-y)*log(1-p)) # we turned products into sums, computationally easier
}

```

Prior

```{r}

logprior <- function(x){dnorm(x, mean = 0, sd = 1, log = TRUE)
}

```


```{r}


for(iter in 1:n_iter){

# update b
  for(j in 1:J){
 # Propose new difficulty
b_prop <- rnorm(1, mean = b_mh[j],sd = prop_b)
# Current log-posterior
logpost_current <-loglik_item( b = b_mh[j],delta = delta_mh[j],y = Y[,j],theta = theta,group = group) +logprior(b_mh[j])
# Proposed log-posterior
logpost_prop <-loglik_item(b = b_prop, delta = delta_mh[j],y = Y[,j],theta = theta,group = group) +logprior(b_prop)
# Log acceptance ratio
log_alpha <-logpost_prop -logpost_current
# Accept or reject
if(log(runif(1)) < log_alpha){
b_mh[j] <- b_prop
acc_b[j] <- acc_b[j] + 1
}

  }

  # update DIF param.


  for(j in 1:J){
# Propose new DIF parameter
delta_prop <- rnorm(1, mean = delta_mh[j],sd = prop_delta)
# Current log-posterior
logpost_current <-loglik_item(b = b_mh[j],delta = delta_mh[j],y = Y[,j],theta = theta,group = group) +
      logprior(delta_mh[j])

# Proposed log-posterior
logpost_prop <-loglik_item(b = b_mh[j],delta = delta_prop,y = Y[,j],theta = theta,group = group) +logprior(delta_prop)

# Log acceptance ratio
log_alpha <-logpost_prop -logpost_current
# Accept or reject
if(log(runif(1)) < log_alpha){

delta_mh[j] <- delta_prop

acc_delta[j] <- acc_delta[j] + 1

    }

  }

  # store samples
 

  b_store[iter, ] <- b_mh

  delta_store[iter, ] <- delta_mh

}

```
Acceptance Rates
```{r}
acc_rate_b <- acc_b / n_iter
acc_rate_delta <- acc_delta / n_iter

acc_rate_b
acc_rate_delta


```

Posterior Summaries

```{r}

delta_post <- delta_store[ (burnin + 1):n_iter,]


delta_mean <- apply(
  delta_post,
  2,
  mean
)

```

CI 

```{r}

delta_ci <- apply(
  delta_post,
  2,
  quantile,
  probs = c(0.025, 0.975)
)

```


Results

```{r}

results <- data.frame(

  Item = 1:J,

  True_DIF = delta,

  Posterior_Mean = round(
    delta_mean,
    3
  ),

  Lower95 = round(
    delta_ci[1,],
    3
  ),

  Upper95 = round(
    delta_ci[2,],
    3
  )

)

results

```

Trace plots
```{r}

pdf(
  "../Figures/mh_trace_plots.pdf",
  width = 9,
  height = 7
)

par(mfrow = c(2, 2))

plot(
  delta_store[, 3],
  type = "l",
  main = expression(delta[3]),
  xlab = "Iteration",
  ylab = "DIF parameter"
)

plot(
  delta_store[, 8],
  type = "l",
  main = expression(delta[8]),
  xlab = "Iteration",
  ylab = "DIF parameter"
)

plot(
  delta_store[, 1],
  type = "l",
  main = expression(delta[1]),
  xlab = "Iteration",
  ylab = "DIF parameter"
)

plot(
  delta_store[, 2],
  type = "l",
  main = expression(delta[2]),
  xlab = "Iteration",
  ylab = "DIF parameter"
)

dev.off()
```




Some tables for the results 

```{r}
# JAGS estimates of delta

stats <- summary(samples)$statistics

delta_rows <- grep("^delta", rownames(stats))

delta_est <- stats[delta_rows, "Mean"]

delta_est

```

table
```{r}


comparison_table <- data.frame(

  Item = 1:J,

  True_DIF = delta,

  MH = round(delta_mean, 3),

  JAGS = round(delta_est, 3)

)

comparison_table

```

plot 
```{r}
pdf(
  "../Figures/mh_vs_jags_estimates.pdf",
  width = 7,
  height = 6
)

plot(
  delta_est,
  delta_mean,
  pch = 19,
  xlab = "JAGS estimate",
  ylab = "MH estimate",
  main = "Metropolis–Hastings versus JAGS"
)

abline(
  a = 0,
  b = 1,
  lty = 2
)

text(
  delta_est,
  delta_mean,
  labels = 1:J,
  pos = 3
)

dev.off()

```
Acceptance rate

```{r}

acceptance_table <- data.frame(

  Item = 1:J,

  Difficulty = round(acc_rate_b,3),

  DIF = round(acc_rate_delta,3)

)

library(knitr)

kable(
  acceptance_table,
  caption = "Acceptance rates of the Metropolis--Hastings algorithm."
)


```


autocorrelation plots

```{r}

pdf(
  "../Figures/mh_autocorrelation_plots.pdf",
  width = 9,
  height = 5
)

par(mfrow = c(1, 2))

acf(
  delta_post[, 3],
  main = expression(delta[3]),
  xlab = "Lag"
)

acf(
  delta_post[, 8],
  main = expression(delta[8]),
  xlab = "Lag"
)

dev.off()

```

acceptance ratio

```{r}


acceptance_table <- data.frame(

  Item = 1:J,

  Difficulty = round(acc_rate_b,3),

  DIF = round(acc_rate_delta,3)

)

acceptance_table

```

Comparing with JAGS

```{r}
comparison_table <- data.frame(

  Item = 1:J,

  True_DIF = delta,

  MH = round(delta_mean,3),

  JAGS = round(delta_est,3)

)

comparison_table


```

### Practical usefulness

the method is validated, now is the time to apply it





#### Empirical data 

```{r}
library(difR)

data(package = "difR")

```

```{r}


data(verbal)
```

```{r}

str(verbal)
names(verbal)
```

response matrix

```{r}
Y_verbal <- as.matrix(verbal[,1:24])

dim(Y_verbal)
```

Group variable 

```{r}

table(verbal$Gender)

group_verbal <- verbal$Gender
```

```{r}

round(colMeans(Y_verbal),3)

```

Classical DIF

```{r}
library(difR)

mh_verbal <- difMH(
  Data = Y_verbal,
  group = group_verbal,
  focal.name = 1
)

mh_verbal$DIFitems

```

```{r}
mh_table <- data.frame(
  Item = 1:24,
  p_value = round(mh_verbal$p.value,4),
  alphaMH = round(mh_verbal$alphaMH,3)
)

mh_table


```


Ability estimates (proxy)


```{r}
theta_verbal <- scale(rowSums(Y_verbal))

```


MH sampler
Follow the same as before, the main thing I did now is that I have adjusted J into J_verbal

```{r}
set.seed(123)

n_iter <- 10000
burnin <- 2000

prop_b <- 0.3
prop_delta <- 0.3

N_verbal <- nrow(Y_verbal)
J_verbal <- ncol(Y_verbal)

```

```{r}
b_mh <- rep(0, J_verbal)
delta_mh <- rep(0, J_verbal)

```

```{r}

b_store <- matrix(NA, n_iter, J_verbal)
delta_store <- matrix(NA, n_iter, J_verbal)

acc_b <- rep(0, J_verbal)
acc_delta <- rep(0, J_verbal)

```

Log likelihood

```{r}
loglik_item <- function(b, delta, y, theta, group){

  eta <- theta - b - delta*group

  p <- plogis(eta)

  sum(y*log(p) +(1-y)*log(1-p))

}

```

Prior Distribution

```{r}
logprior <- function(x){

  dnorm(
    x,
    mean = 0,
    sd = 1,
    log = TRUE
  )

}

```

Metropolis-Hastings

```{r}

for(iter in 1:n_iter){
  for(j in 1:J_verbal){

    b_prop <- rnorm(1, mean = b_mh[j], sd = prop_b)
logpost_current <-loglik_item(b_mh[j],delta_mh[j],Y_verbal[,j],theta_verbal,
        group_verbal) +
      logprior(b_mh[j])

    logpost_prop <-
      loglik_item(b_prop,delta_mh[j],Y_verbal[,j],theta_verbal,group_verbal
      ) +
      logprior(b_prop)

    if(
      log(runif(1)) <
      (logpost_prop - logpost_current)
    ){

      b_mh[j] <- b_prop

      acc_b[j] <- acc_b[j] + 1

    }

  }


  # for DIF 

  for(j in 1:J_verbal){

    delta_prop <- rnorm(1,mean = delta_mh[j],sd = prop_delta)

    logpost_current <-loglik_item(b_mh[j],delta_mh[j],Y_verbal[,j],theta_verbal,group_verbal
      ) +
      logprior(
        delta_mh[j]
      )

    logpost_prop <-
      loglik_item(
        b_mh[j],
        delta_prop,
        Y_verbal[,j],
        theta_verbal,
        group_verbal
      ) +
      logprior(
        delta_prop
      )

    if(
      log(runif(1)) <
      (logpost_prop - logpost_current)
    ){

      delta_mh[j] <- delta_prop

      acc_delta[j] <- acc_delta[j] + 1

    }

  }



  b_store[iter,] <- b_mh

  delta_store[iter,] <- delta_mh

}


```


Acceptance rate

```{r}

acc_rate_b <- acc_b / n_iter
acc_rate_delta <- acc_delta / n_iter

acc_rate_b
acc_rate_delta
```

Posterior summaries

```{r}
delta_post <- delta_store[
  (burnin+1):n_iter,
]

```

Posterior mean

```{r}
delta_mean <- apply(
  delta_post,
  2,
  mean
)

```

CI

```{r}

delta_ci <- apply(
  delta_post,
  2,
  quantile, 
  probs = c(0.025, 0.975)
)

dim(delta_ci)
```

Results table

```{r}
results_verbal <- data.frame(

  Item = 1:J_verbal,

  Posterior_Mean =
    round(delta_mean,3),

  Lower95 =
    round(delta_ci[1,],3),

  Upper95 =
    round(delta_ci[2,],3)

)

results_verbal

```
Trace plots

```{r}
par(mfrow = c(2,1))

plot(
  delta_store[,1],
  type="l",
  main=expression(delta[1])
)

plot(
  delta_store[,2],
  type="l",
  main=expression(delta[2])
)

plot(
  delta_store[,3],
  type="l",
  main=expression(delta[3])
)

plot(
  delta_store[,4],
  type="l",
  main=expression(delta[4])
)

```

Combined table 

```{r}

combined_results <- data.frame(

  Item = 1:J_verbal,

  MH_p_value = round(mh_verbal$p.value, 4),

  MH_alpha = round(mh_verbal$alphaMH, 3),

  Posterior_Mean = round(delta_mean, 3),

  Lower95 = round(delta_ci[1,], 3),

  Upper95 = round(delta_ci[2,], 3)

)

combined_results$Bayesian_DIF <-
  ifelse(
    combined_results$Lower95 > 0 |
    combined_results$Upper95 < 0,
    "Yes",
    "No"
  )

combined_results$MH_DIF <-
  ifelse(
    combined_results$MH_p_value < .05,
    "Yes",
    "No"
  )

combined_results

write.csv(
  combined_results,
  "../Data/processed/empirical_dif_comparison.csv",
  row.names = FALSE
)

```

extract the tables for overleaf

```{r}

library(xtable)

print(
  xtable(combined_results,
         caption = "Comparison of Mantel--Haenszel and Bayesian DIF results.",
         label = "tab:empirical"),
  include.rownames = FALSE
)

```

#### hypothesis testing 


```{r}
library(bain)

library(restriktor)

```


```{r}
dat_goric <- data.frame(

d6  = delta_post[,6],

d16 = delta_post[,16],

d17 = delta_post[,17],

d19 = delta_post[,19]

)


```


```{r}

fit_goric <- lm(
cbind(d6,d16,d17,d19) ~ 1,
data = dat_goric
)


```




```{r}
dat <- data.frame(
  d6 = delta_post[,6],
  d16 = delta_post[,16],
  d17 = delta_post[,17],
  d19 = delta_post[,19]
)

fit <- lm(cbind(d6,d16,d17,d19) ~ 1, data = dat)

```





```{r}

d6 <- delta_post[,6]

d16 <- delta_post[, 16]

d17 <- delta_post[, 17]

d19 <- delta_post[, 19]

```

measurement invariance

```{r}

H1 <- mean(
  abs(d6)  < 0.10 &
  abs(d16) < 0.10 &
  abs(d17) < 0.10 &
  abs(d19) < 0.10
)

H1

```
Expected directions
```{r}

H2 <- mean(
  d6  > 0 &
  d16 < 0 &
  d17 < 0 &
  d19 < 0
)

H2

```
Ordered effects

```{r}
H3 <- mean(
  d16 < d17 &
  d17 < d19 &
  d19 < 0 &
  d6 > 0
)

H3



```

Summarise the results

```{r}

hypothesis_results <- data.frame(

  Hypothesis = c(
    "Measurement invariance",
    "Directional DIF",
    "Ordered DIF effects"
  ),

  Posterior_Probability = round(
    c(H1, H2, H3),
    3
  )

)

hypothesis_results

```

Fit log regression for each item 

```{r}
fit6 <- glm(
  Y_verbal[,6] ~ theta_verbal + group_verbal,
  family = binomial
)

fit16 <- glm(
  Y_verbal[,16] ~ theta_verbal + group_verbal,
  family = binomial
)

fit17 <- glm(
  Y_verbal[,17] ~ theta_verbal + group_verbal,
  family = binomial
)

fit19 <- glm(
  Y_verbal[,19] ~ theta_verbal + group_verbal,
  family = binomial
)



```

```{r}

bain6 <- bain(
  fit6,
  "group_verbal = 0;
   group_verbal > 0;
   group_verbal < 0"
)

summary(bain6)


```

```{r}

bain6$fit

```


```{r}

bain16 <- bain(
  fit16,
  "group_verbal = 0;
   group_verbal < 0;
   group_verbal > 0"
)

bain17 <- bain(
  fit17,
  "group_verbal = 0;
   group_verbal < 0;
   group_verbal > 0"
)

bain19 <- bain(
  fit19,
  "group_verbal = 0;
   group_verbal < 0;
   group_verbal > 0"
)



summary(bain16)
summary(bain17)
summary(bain19)
```


full table

```{r}

hypothesis_results <- data.frame(

  Item = c(6,16,17,19),

  No_DIF = c(
    bain6$fit["H1","Fit_eq"],
    bain16$fit["H1","Fit_eq"],
    bain17$fit["H1","Fit_eq"],
    bain19$fit["H1","Fit_eq"]
  ),

  Expected_Direction = c(
    bain6$fit["H2","Fit_in"],
    bain16$fit["H2","Fit_in"],
    bain17$fit["H2","Fit_in"],
    bain19$fit["H2","Fit_in"]
  ),

  Opposite_Direction = c(
    bain6$fit["H3","Fit_in"],
    bain16$fit["H3","Fit_in"],
    bain17$fit["H3","Fit_in"],
    bain19$fit["H3","Fit_in"]
  )

)

round(hypothesis_results,3)


```


```{r}
hypothesis_results <- data.frame(

  Item = c(6,16,17,19),

  No_DIF = c(
    bain6$fit["H1","Fit_eq"],
    bain16$fit["H1","Fit_eq"],
    bain17$fit["H1","Fit_eq"],
    bain19$fit["H1","Fit_eq"]
  ),

  Expected_DIF = c(
    bain6$fit["H2","Fit_in"],
    bain16$fit["H2","Fit_in"],
    bain17$fit["H2","Fit_in"],
    bain19$fit["H2","Fit_in"]
  ),

  Opposite_DIF = c(
    bain6$fit["H3","Fit_in"],
    bain16$fit["H3","Fit_in"],
    bain17$fit["H3","Fit_in"],
    bain19$fit["H3","Fit_in"]
  )

)

round(hypothesis_results,3)


```

Using PMPb

```{r}
hypothesis_results <- data.frame(

Item = c(6,16,17,19),

No_DIF = c(
bain6$fit["H1","PMPb"],
bain16$fit["H1","PMPb"],
bain17$fit["H1","PMPb"],
bain19$fit["H1","PMPb"]
),

Expected_DIF = c(
bain6$fit["H2","PMPb"],
bain16$fit["H2","PMPb"],
bain17$fit["H2","PMPb"],
bain19$fit["H2","PMPb"]
),

Opposite_DIF = c(
bain6$fit["H3","PMPb"],
bain16$fit["H3","PMPb"],
bain17$fit["H3","PMPb"],
bain19$fit["H3","PMPb"]
)

)

round(hypothesis_results,3)

```

```{r}

hypothesis_results <- data.frame(

Item = c(6,16,17,19),

No_DIF = c(
bain6$fit["H1","PMPb"],
bain16$fit["H1","PMPb"],
bain17$fit["H1","PMPb"],
bain19$fit["H1","PMPb"]
),

Expected_DIF = c(
bain6$fit["H2","PMPb"],
bain16$fit["H2","PMPb"],
bain17$fit["H2","PMPb"],
bain19$fit["H2","PMPb"]
),

Opposite_DIF = c(
bain6$fit["H3","PMPb"],
bain16$fit["H3","PMPb"],
bain17$fit["H3","PMPb"],
bain19$fit["H3","PMPb"]
),

Unconstrained = c(
bain6$fit["Hu","PMPb"],
bain16$fit["Hu","PMPb"],
bain17$fit["Hu","PMPb"],
bain19$fit["Hu","PMPb"]
)

)

round(hypothesis_results,3)


```

Table for overleaf

```{r}

library(xtable)

print(
xtable(
hypothesis_results,
caption = "Posterior model probabilities obtained from Bayesian hypothesis testing.",
label = "tab:bain"
),
include.rownames = FALSE
)

```

GORIC


```{r}

beta <- c(
  delta_mean[6],
  delta_mean[16],
  delta_mean[17],
  delta_mean[19]
)

names(beta) <- c("d6","d16","d17","d19")

beta

```


```{r}

Sigma <- cov(
  delta_post[, c(6,16,17,19)]
)

Sigma

```


```{r}

H1 <- "d6 = d16 = d17 = d19 = 0"
H2 <- "
d6 > 0
d16 < 0
d17 < 0
d19 < 0
"
H3 <- "
d16 < d17
d17 < d19
d19 < 0
d6 > 0
"
H4 <- "
d16 < d17
d16 < d19
d6 > 0
"




```

